Modeling electrostatics in molecular biology: A tutorial of DelPhi and associated resources [Article v\versionnumber]

### 2019-09-16 article-manuscript/article-manuscript-tutorial-v01.tex
English corrections: Revision-1

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\graphicspath{{figures/}}

@@

\textit{Visualization of the surface and its potentials}:
The \texttt{zphi} file with the potential and coordinate information can be used for visualizing the surface and the distribution of the potentials on it. It can also be used for other statistical analysis a user might be interested in. To visualize the surface on VMD\cite{humphrey1996vmd}, a TCL script is available for download from \url{http://common/compbio/delphi}. The name of the downloaded file \texttt{seeSurface.tcl}

@@

\subsubsection{Computing electrostatic potential, energy of interaction and forces between two sets of atoms}
In this section, we will describe how to calculate the electrostatic potential, energy of interaction, and force between two sets of atoms using ‘\texttt{frc}’ option. Scheme is shown in Figure \ref{fig:elec_force}.

@@

This will output \ttvar{phimap1.cube} which is then read by the second “child” run. The acenter to be set to the center of Asp39 of barstar in the child run parameter file. The child run reads the \ttvar{phimap1.cube} and outputs \ttvar{phimap2.cube}. The resolution of the child run is higher than in the parent run (\texttt{scale=4.0}) and the calculation box is smaller; the \texttt{acenter} option is used to indicate where is the center of the child box; the boundary condition is set as 3 to indicate the boundary values are calculated based on the input \texttt{phimap1.cube}; an frc option is used to calculate the potential and field on residue Asp39 from barstar. 
The child run can be done as below.

@@

\textbf{Electrostatic component of binding energy Gaussian approach}.
In the Gaussian approach there is not a strict solvent-solute boundary which separates the mediums of two different dielectric, rather the regions in solute and solvent are described by a smooth dielectric function. Because of this fact, energy partitioning described in previous paragraph is not applicable. Therefore, the grid energy of the system which implicitly accounts Coulombic and polar component of the solvation energy is used in this case. However, to be able to obtain electrostatic component of binding, we must use the same point as grid center and same number of grid sizes and scale for complex and monomers. To keep the same point as grid center, one can compute the center of complex and use \ttvar{acenter} input command as \texttt{acenter(xc, yc, zc)} (where \ttvar{xc}, \ttvar{yc}, \ttvar{zc} are the x, y, z coordinates of center for the complex) in all the three calculations. In addition, we are required to activate the Gaussian module by providing \texttt{gaussian=1} in the parameter and the \texttt{sigma=0.65} parameter for the gaussian.
We can add the following set of parameters and their respective values to the parameter files for Gaussian run using Delphi for present example.
\begin{verbatim}
acent(28.459,36.960,8.674)
gaussian=1
sigma=0.65
srfcut=20
\end{verbatim}
\ttvar{Example_3.1.6/} and \ttvar{Example_3.1.6/gaussian/} directories, change current working directory to \ttvar{Example_3.1.6/gaussian/}. Now we can do the Delphi run using following command.
\begin{verbatim}
$DELPHI_EXE param_barnase-barstar_gauss.prm \
                   > param_barnase-barstar_gauss.log
$DELPHI_EXE param_barnase_gauss.prm \
                   > param_barnase_gauss.log
$DELPHI_EXE param_barstar_gauss.prm \
                   > param_barstar_gauss.log
\end{verbatim}
The total grid energies for barnase-barstar complex, barnase and barstar are found from the respective .dat files and the electrostatic component of binding is computed as below via Method-1. 15975.45 - (8825.55 + 7165.57) kT = -15.66 kT. In the present case, note that the electrostatic component of binding energy is very different from the energy calculated with traditional two-dielectric approach. This is due to different dielectric distribution applied in these two approaches. One should use the traditional two-dielectric if modeling involves rigid structures (energy minimized or snapshots from MD simulations). The Gaussian approach is applicable for cases when one wants to use single structure to generate ensemble average quantity (see \cite{chakravorty2018gaussian} for more details).

@@

In DelPhi basic example 9, the salt dependence is calculated using binding free energy under two different salt concentrations, 0 and 0.15M. Usage of grid energy is recommended when calculating salt dependence. The output from DelPhi should be as following,

@@

DelPhiForce\cite{li2017delphiforce} is developed and made available within DelPhi distribution to calculate electric field, forces, and energy for a two-molecule system. It can be download from the DelPhi webpage at \url{http:// compbio.clemson.edu/downloadDir/delphiforce.tar.gz}. The DelPhiForce web server is available at \url{http://compbio.clemson.edu/delphi-force/}. Using DelPhiForce output files, the electrostatic forces can be visualized in VMD\cite{humphrey1996vmd}. Once \texttt{DelPhiForce.tar.gz} is downloaded, In the \texttt{DelPhiForce/bin} folder, there are several files e.g. 
\texttt{DelPhiForrce.sh}, \texttt{ForceGen.sh}, \texttt{draw\_arrow.tcl} which will be used in this example.

The two input files required to run DelPhiForce are reference molecule (file1) and probe molecule (file2), both of which are in \texttt{.pqr} format. To run the DelPhiForce, go to the working folder, copy all files from the DelPhiForce/bin/ folder to it, then run DelPhiForce using command below:

\begin{verbatim}
./DelPhiForce.sh file1 file2 out
\end{verbatim}

@@

In order to visualize the forces, first launch VMD, open file2. Then, in the VMD Tk Console, run the commands:

@@

Then open the file1 in VMD. The electrostatic force will be displayed.  Each arrow illustrates an electrostatic force on a residue of probe molecule. The tail of an arrow is the mass center of the corresponding residue. The length of an arrow indicates the magnitude of the force.

\begin{figure}[hbt!]
\includegraphics[width=\linewidth]{Figure_14.png}
\caption{Electrostatic forces shown with blue arrows on residues of dynein computed by DelPhiForce. For clarity the dynein is positioned 15 $\text{\AA}$ away from it's bound position.}
\label{fig:delphiforce}
\end{figure}

@@

\item Diffusion constant of the simulation: In the simulation, Langevin dynamics is applied for the modeling of dynamics in the molecular system, where the diffusion effects are simulated via random applied forces and velocity dependent frictions. We provide users with an option to adjust the damping coefficient if needed.

\item Simulation mode: Based on type of simulation users are provided with three different simulation modes: Accelerated mode, such that the atomic DefPhiForce are calculated constantly until the end of the simulation; Cut-off mode, such that the atomic DelPhiForces are calculated only for ligand atoms far away from the receptor than the cut-off distance in MD simulations ; Scaled mode, such that atomic DelPhiForces are corrected for force calculated via GB, which guarantees that there is no double counting of electrostatic force in MD.

@@

\subsubsection{SAAMBE}
The Single Amino Acid Mutation based change in Binding free Energy (SAAMBE)\cite{petukh2015predicting} web-server predicts the change of protein binding free energy due to a mutation. The web-server\cite{petukh2015predicting,petukh2016saambe} can be accessed via \url{http://compbio.clemson.edu/saambe_web-server/}. 
The user input section is categorized into three sections:
\begin{enumerate}
    \item PDB selection: In this section, user need to provide a structure file by uploading a PDB file from their local file system.

    \item Partner selection: There are two inputs (Partner 1 and Partner 2) in this section. User must supply the chain of partner 1 and chain of partner 2 respectively as the input. If multiple chains construct partner 1 or partner 2, user needs to provide all the chains. For example, if chain A and chain B construct partner 1, then user must mention AB in place of Partner 1.

    \item Mutation: This section contains four sub sections:
    \begin{enumerate}
	    \item Position: position of the particular residue i.e. the residue number, which is replaced in the mutant, need to be provided in the box.

@@

\subsubsection{SAAFEC} Single Amino Acid Folding free Energy Changes (SAAFEC) program\cite{getov2016saafec} predicts the folding free energy changes due to missense mutation. SAAFEC is available as a web server and accessible via http://compbio.clemson.edu/SAAFEC/. SAAFEC runs on Clemson University’s Palmetto Supercomputer cluster. Once user click the Use SAAFEC button, it would direct to a page containing Input Parameter Form. To run a job in SAAFEC, user needs to provide following four job input parameters:
\begin{enumerate}
\item \textit{Location of Residue}: Residue number at which mutation takes place.
\item \textit{Initial Residue (One letter abbreviation)}: Amino acid type of wild type residue.
\item \textit{Mutated Residue (One letter abbreviation)}: Amino acid type of mutant. 
\item \textit{Chain}: The chain ID where the mutation occurs.
\end{enumerate}
As a final parameter, user must upload the PDB file of the desired structure in the Upload your PDB file box. On clicking the Next button, it would direct user to a page that includes File upload and input parameter form for SAAFEC run. The page shows all the Input Details that user already provided. The page also provides a link that directs to the result page. If the job is still running, user will be informed by showing ``The SAAFEC job is still being executed''. Once the job is completed and if all the user-provided inputs are correct, the page shows the Output Results. User can also download the same information by clicking Download Result File. Overall, following three downloadable result files are generated after completion of a successful job:

@@

DelPhiPKa\cite{wang2015pka} is a DelPhi based C++ program, which predicts the pKa’s of ionizable and polar groups in proteins, DNA and RNAs. The method also takes into account the salt concentration in the medium\cite{pahari2018delphipka}. DelPhiPKa web-server is built on DelPhiPKa program and user can access it at \url{http://compbio.clemson.edu/pka_web-server/}. The web-server requires the following input parameters for a successful run:

>>>>

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage[italic]{mathastext}
\usepackage{hyperref}
\graphicspath{{figures/}}

@@

\textit{Visualization of the surface and its potentials}:
The \texttt{zphi} file with the potential and coordinate information can be used for visualizing the surface and the distribution of the potentials on it. It can also be used for other statistical analysis a user might be interested in. To visualize the surface on VMD\cite{humphrey1996vmd}, a TCL script is available for download from \url{http://compbio.clemson.edu/downloadDir/seeSurface.tcl}. The name of the downloaded file \texttt{seeSurface.tcl}

@@

\subsubsection{Computing electrostatic potential, energy of interaction and forces between two sets of atoms}
In this section, we will describe how to calculate the electrostatic potential, energy of interaction and force between two sets of atoms using ‘\texttt{frc}’ option. Scheme is shown in Figure \ref{fig:elec_force}.

@@

This will output \ttvar{phimap1.cube} which is then read by the second “child” run. The acenter to be set to the center of Asp39 of barstar in the child run parameter file. The child run reads the \ttvar{phimap1.cube} and outputs \ttvar{phimap2.cube}. The resolution of the child run is higher than in the parent run (\texttt{scale=4.0}) and the calculation box is smaller; the \texttt{acenter} option is used to indicate where is the center of the child box; the boundary condition is set as 3 to indicate the boundary values are calculated based on the input \texttt{phimap1.cube}; an frc option is used to calculate the potential and field on residue Asp39 of barstar. 
The child run can be done as below.

@@

\textbf{Electrostatic component of binding energy Gaussian approach}.
In the Gaussian approach there is no strict solvent-solute boundary which separates the mediums of two different dielectric, rather the regions in solute and solvent are described with smooth dielectric function. Because of this fact, energy partitioning described in previous paragraph is not applicable. Therefore, the grid energy of the system which implicitly accounts Coulombic and polar component of the solvation energy is used in this case. However, to be able to obtain electrostatic component of binding, we must use the same point as grid center and same number of grid sizes and scale for complex and monomers. To keep the same point as grid center, one can compute the center of complex and use \ttvar{acenter} input command as \texttt{acenter(xc, yc, zc)} (where \ttvar{xc}, \ttvar{yc}, \ttvar{zc} are the x, y, z coordinates of center for the complex) in all the three calculations. In addition, we are required to activate the Gaussian module by providing \texttt{gaussian=1} in the parameter and the \texttt{sigma=0.65} parameter for the gaussian.
We can add the following set of parameters and their respective values to the parameter files for Gaussian run using Delphi for present example.
\begin{verbatim}
acent(28.459,36.960,8.674)
gaussian=1
sigma=0.65
srfcut=20
\end{verbatim}
\ttvar{Example_3.1.6/} and \ttvar{Example_3.1.6/gaussian/} directories, change current working directory to \ttvar{Example_3.1.6/gaussian/}. Now we can do the Delphi run using following command.
\begin{verbatim}
$DELPHI_EXE param_barnase-barstar_gauss.prm \
                   > param_barnase-barstar_gauss.log
$DELPHI_EXE param_barnase_gauss.prm \
                   > param_barnase_gauss.log
$DELPHI_EXE param_barstar_gauss.prm \
                   > param_barstar_gauss.log
\end{verbatim}
The total grid energies for barnase-barstar complex, barnase and barstar are found from the respective .dat files and the electrostatic component of binding is computed as below via Method-1. 15975.45 - (8825.55 + 7165.57) kT = -15.67 kT. In the present case, note that the electrostatic component of binding energy is very different from the energy calculated with traditional two-dielectric approach. This is due to different dielectric distribution applied in these two approaches. One should use the traditional two-dielectric if modeling involves rigid structures (energy minimized or snapshots from MD simulations). The Gaussian approach is applicable for cases when one wants to use single structure to generate ensemble average quantity (see \cite{chakravorty2018gaussian} for more details).

@@

In the \texttt{Example\_3.1.8}, the salt dependence is calculated using binding free energy under two different salt concentrations, 0 and 0.15M. Usage of grid energy is recommended when calculating salt dependence. One can do DelPhi calculations at varying salt concentration by executing the provided bash script \ttvar{run.sh} as given below:

\begin{verbatim}
bash run.sh
\end{verbatim}

The output from DelPhi should be as following,

@@

DelPhiForce\cite{li2017delphiforce} is developed and made available within DelPhi distribution to calculate electric field, forces, and energy for a two-molecule system. It can be downloaded from the DelPhi webpage at \url{http:// compbio.clemson.edu/downloadDir/delphiforce.tar.gz}. The DelPhiForce web server is available at \url{http://compbio.clemson.edu/delphi-force/}. Using DelPhiForce output files, the electrostatic forces can be visualized in VMD\cite{humphrey1996vmd}. Once \texttt{DelPhiForce.tar.gz} is downloaded. In the \texttt{DelPhiForce/bin} folder, there are several files e.g. 
\texttt{DelPhiForrce.sh}, \texttt{ForceGen.sh}, \texttt{draw\_arrow.tcl} which will be used in this example.

The two input files required to run DelPhiForce are reference molecule (\ttvar{file1}) and probe molecule (\ttvar{file2}), both of which are in \texttt{.pqr} format. To run the DelPhiForce, go to the example folder, copy all files from the DelPhiForce/bin/ folder to it, then run DelPhiForce using command below:

\begin{verbatim}
./DelPhiForce.sh -1 file1 -2 file2 \
    -e absolute_path_to/delphicpp -o out
\end{verbatim}

@@

In order to visualize the forces, first launch VMD, open \ttvar{file2}. Then, in the VMD Tk Console, run the commands:

@@

Then open the \ttvar{file1} in VMD. The electrostatic force will be displayed.  Each arrow illustrates an electrostatic force on a residue of probe molecule. The tail of an arrow is the mass center of the corresponding residue.

\begin{figure}[hbt!]
\includegraphics[width=\linewidth]{Figure_14.png}
\caption{Electrostatic forces generated by tubulin dimer (cyan colored cartoon representation) on to kinesin microtubule binding domain (line representation), placed 15 $\text{\AA}$ away from its bound position. Orange arrows represent forces acting on each residue, while the green arrow shows total resultant force. The length of arrow reflects the calculated magnitude of the force.}
\label{fig:delphiforce}
\end{figure}

@@

\item Diffusion constant of the simulation: In the simulation, Langevin dynamics is applied for the modeling of dynamics of the molecular system, where the diffusion effects are simulated via randomly applied forces and velocity dependent frictions. We provide users with an option to adjust the damping coefficient if needed.

\item Simulation mode: Based on type of simulation users are provided with three different simulation modes: Accelerated mode, such that the atomic DefPhiForce are calculated constantly until the end of the simulation; Cut-off mode, such that the atomic DelPhiForces are calculated only for ligand atoms far away from the receptor than the cut-off distance in MD simulations ; Scaled mode, such that the atomic DelPhiForces are corrected for force calculated via GB, which guarantees that there is no double counting of electrostatic force in MD.

@@

\subsubsection{SAAMBE}
The Single Amino Acid Mutation based change in Binding free Energy (SAAMBE)\cite{petukh2015predicting} web-server predicts the change of protein binding free energy due to a mutation. The web-server\cite{petukh2015predicting,petukh2016saambe} can be accessed via \url{http://compbio.clemson.edu/saambe_webserver/}. 
The user input section is categorized into three sections:
\begin{enumerate}
    \item PDB selection: In this section, user need to provide a structure file by uploading a PDB file from their local file system.

    \item Partner selection: There are two inputs (Partner 1 and Partner 2) in this section. User must supply the chain of partner 1 and chain of partner 2 respectively as the input. If multiple chains construct partner 1 or partner 2, user needs to provide all the chains. For example, if chain A and chain B construct partner 1, then user must provide AB in place of Partner 1.

    \item Mutation: This section contains four sub sections:
    \begin{enumerate}
	    \item Position: position of the particular residue, i.e. the residue number, which is replaced in the mutant, need to be provided in the box.

@@

\subsubsection{SAAFEC} Single Amino Acid Folding free Energy Changes (SAAFEC) program\cite{getov2016saafec} predicts the folding free energy changes due to missense mutation. SAAFEC is available as a web server and accessible via \url{http://compbio.clemson.edu/SAAFEC/}. SAAFEC runs on Clemson University’s Palmetto Supercomputer cluster. Once user click the Use SAAFEC button, it would direct to a page containing Input Parameter Form. To run a job in SAAFEC, user needs to provide following four job input parameters:
\begin{enumerate}
\item \textit{Location of Residue}: Residue number at which mutation takes place.
\item \textit{Initial Residue (One letter abbreviation)}: Amino acid type of wild type residue.
\item \textit{Mutated Residue (One letter abbreviation)}: Amino acid type of mutant. 
\item \textit{Chain}: The chain ID where the mutation occurs.
\end{enumerate}
As a final parameter, user must upload the PDB file of the desired structure in the Upload your PDB file box. On clicking the Next button, it would direct user to a page that includes File upload and input parameter form for SAAFEC run. The page shows all the Input Details that user already provided. The page also provides a link that directs to the result page. If the job is still running, user will be informed by showing ``The SAAFEC job is still being executed''. Once the job is completed and if all the user-provided inputs are correct, the page shows the Output Results. User can also download the same information by clicking Download Result File. Overall, following three downloadable result files are generated after completion of a successful job:

@@

DelPhiPKa\cite{wang2015pka} is a DelPhi based C++ program, which predicts the pKa’s of ionizable and polar groups in proteins, DNA and RNAs. The method also takes into account the salt concentration in the medium\cite{pahari2018delphipka}. DelPhiPKa web-server is built on DelPhiPKa program and user can access it at \url{http://compbio.clemson.edu/pka_webserver/}. The web-server requires the following input parameters for a successful run:
